---
title: "Introduction to R"
output:
  github_document: default
  html_document: default
  pdf_document: default
---


### R as a calculating environment

R can be used as a powerful calculator.
For arithmetic calculations:
```{r}
(1+1/100)^100
```
```{r}
17%%5
```
```{r}
17%/%5
```
[1] implies this is item 1 in a vector of output.


symbol | meaning
------ | --------
+ | addition
- | subtraction
* | multiplication
/ | division
^ | exponential
%% | modulus
%/% | integer division

### Built-in function in R

R has a number of built in functions:  
sin(x), cos(x), tan(x), exp(x), log(x), sqrt(x), floor(x), ceiling(x), round(x), ???

```{r}
exp(1)
options(digits = 16)
exp(1)
pi
sin(pi/6)
```

### Variable

* We can assign a value to a variable and use the variable.

* For the assignment, we use command <-

* Variable names made up of letters, numbers, . or _

  * provided it starts with a letter, or . then a letter.
  * names are case sensitive.
  * for example,
    * x, y, my_variable, a1, a2, .important_variable, x.input
  * wrong name:
    * 2016_income, .1grade, _x, y@gmail.com
    
To display the value of a variable x, we type x 
* or print(x) or show(x).
```{r}
x <- 100
x

print(x)

show(x)
```

We can show the outcome of assignment by parentheses.
```{r}
(y <- (1+1/x)^x)
```

When assigning, the right-hand side is evaluated first, then that value is placed in the variable on the left-hand side.
```{r}
n <- 1
n <- n+1
n
```

R allows the use of = for variable assignment, in common with most programming languages.

### Functions

Takes one or more argument (inputs) and produces one or more outputs (return values).

```{r}
seq(from = 1, to = 9, by =2)
```
```{r}
seq(from = 1, to = 9)
```

You can access the built-in help by help(function name) or ?function name:

Every function has a default order for arguments.  
If you provide arguments in this order, then they do not need to be named.
```{r}
seq(1, 9, 2)
```
```{r}
seq(to = 9, from = 1)
```
```{r}
seq(by=-2, 9, 1)
```

### Vectors

* indexed list of variables
* the name of the i-th element of vector x is x[i]
* three basic functions for constructing vectors

```{r}
(x <- seq(1, 20, by = 2))
```

```{r}
(y <- rep(3, 4))
```

```{r}
(z <- c(y, x))
```

* another method for sequence
```{r}
(x <- 100:110)
```
```{r}
(y <- 110:100)
```

* vector and index
```{r}
(x<- 100:110)

i <- c(1, 3, 2)

x[i]
```

minus index:
```{r}
j <- c(-1, -2, -3)
x[j]
```

empty vetor:
```{r}
x <- c()
```

elementwise algebraic operation:
```{r}
x <- c(1, 2, 3)
y <- c(4, 5, 6)
x*y
y^x
```

with unequal length of vectors:
```{r}
c(1, 2, 3, 4) + c(1, 2)

2 * c(1, 2, 3)

(1:3)^2
```

This works but with warning message:
```{r, eval=FALSE}
c(1, 2, 3) + c(1, 2)
```

functions taking vectors
```{r}
sqrt(1:3)

mean(1:6)

sort(c(5, 1, 3, 4, 2))
```
#### Example : mean and variance

compare computed mean and variance with built-in functions

```{r}
x <- c(1.2, 0.9, 0.8, 1, 1.2)

x.mean <- sum(x)/length(x)

x.mean - mean(x) 
```
```{r}
x.var <- sum((x-x.mean)^2)/(length(x)-1)
x.var - var(x)
```


Example : simple numerical integration

```{r}
dt <- 0.005
t <- seq(0, pi/6, by = dt)
ft <- cos(t)
(I <- sum(ft) * dt)
```
t is a vector and ft is also a vector.
```{r}
I - sin(pi/6)
```
Note the difference between the numerical integration and theoretical value.


Example : exponential limit
```{r}
x <- seq(10, 200, by = 10)

y <- (1 + 1/x)^x

exp(1) - y
```
```{r}
plot(x, y)
```

### Missing data

in R, missing data is represented by NA.

```{r}
a <- NA   # assign NA to variable A
is.na(a)     # is it missing?
```
```{r}
a <- c(11, NA, 13)
is.na(a)
```
```{r}
mean(a)
```

```{r}
mean(a, na.rm = TRUE) #NAs can be removed
```

### Expression and assignment

Expression is a phrase of code that can be executed.
```{r}
seq(10, 20, by=3)
```

```{r}
4
```
```{r}
mean(c(1,2,3))
```
```{r}
1 > 2
```

If the evaluation of the expression is saved using <-, then it called an assignment.
```{r}
x1 <- seq(10, 20, by=3)
```
```{r}
x2 <- 1>2
```

A logical expression is formed using 
* the comparison operators 
  * <, >, <=, >=, ==, and != (not equal to)
* and the logical operators 
  * & (and), | (or), and ! (not).
The value of a logical expression is either TRUE or FALSE.
* The integers 1 and 0 can also be used as TRUE or FALSE.
```{r}
c(0, 0, 1, 1) | c(0, 1, 0, 1)
```

### x[subset]

We can extract a subvector using a subset as a vector of TRUE/FALSE.

```{r}
x <- 1:10
x%%4 == 0
```
```{r}
( y <- x[ x%%4==0 ] )
```

R also provide subset function, which ignore NA.
* whereas x[subset] preserve NA.
```{r}
x <- c(1, NA, 3, 4)
x[x >2]
```
```{r}
subset(x, subset = x>2 )
```

For the index position of TRUE elements, use which(x)
```{r}
x <- c(1, 1, 2, 3, 5, 8, 13)
which(x%%2 == 0)
```

Example : rounding error

Many floating numbers are subject to rounding errors in digital computers.
```{r}
2*2 == 4
```
```{r}
sqrt(2)*sqrt(2) == 2
```
The solution is to use all.equal(x,y), which returns TRUE if the difference between x and y is smaller than some tolerance.

```{r}
all.equal(sqrt(2)*sqrt(2), 2)
```


### Matrix

Matrix is created from a vector using the function matrix:

* matrix( data, nrow =1, ncol=1, byrow=TRUE )
  * data : vector of length at most nrow*ncol
    * if length of vector< nrow*ncol, then data is reused as many times as is needed
  * nrow : number of rows
  * ncol : number of columns
  * byrow = TRUE : fill the matrix up row-by-row
  * byrow = FALSE : fill the matrix up column-by-column, default

diag(x) : create diagonal matrix
rbind(...) : join matrices with rows of the same length
cbind(...) : join matrices with columns of the same length


Exampe:
```{r}
(A <- matrix(1:6, nrow=2, ncol=3, byrow=TRUE))
```
```{r}
A[1, 3] <- 0
A
```

```{r}
A[, 2:3]
```

```{r}
(B <- diag(c(1,2,3)))
```

Matrix operation
* Usual algebraic operations, including ```*```, act elementwise.
* To perform matrix operation, we use ```%*%```.
* ```nrow(x)```, ```ncol(x)```
* ```det(x)``` : determinant of ```x```
* ```t(x)``` : transpose of ```x```
* ```solve(A, B)``` : returns ```x``` such that ```A %*% x = B```
* If ```A``` is invertable, the ```solve(A)``` is the inverse of ```A```.

```{r}
A <- matrix(c(3,5,2,3), nrow=2, ncol=2)
B <- matrix(c(1,1,0,1), nrow=2, ncol=2)
```
```{r}
A %*% B
```
```{r}
A.inv <- solve(A)
A %*% A.inv   # we observe rounding error
```
```{r}
A^(-1)   #This is not an inverse. ^(-1) applies elementwise.
```


### Workspace

The objects that you create using R remain in existence until you explicitly delete them.

* ```rm(x)``` : remove object x
* ```rm(list=ls())``` : remove all objects


### Working directory

When you run R, it uses one of the directories on your hard drive as a working directory,
* where it looks for user-written programs and data files.

Check the working directory.
```{r, eval=FALSE}
getwd()
```

Change the working directory to "dir"

```{r, eval=FALSE}
setwd("dir")
```

/ is for directory and file address, . refers current directory, .. refers parent directory

### Writing script

We can type and evaluate all possible R expression at the prompt, it is much more convenient to write scripts,
* which simply comprise collections of R expression.
* We use the terms program and code synonymously with script.
You can use built-in editor in Rgui or Rstudio.
* or text-editor like Tinn-R, emacs

### Help

To find out more about an R command or function x, you can type ```help(x)``` or just ```?x```.

If you cannot remember the exact name, then ```help.search("x")```.

HTML help command : ```help.start()```

### package

R provides various useful packages to help you.
https://cran.r-project.org/web/packages/

To install a package:  
```install.packages("packagename")```  
To access the package:  
```library("packagename")```

Or use package menu.

