---
title: "Programming with functions"
output:
  github_document: default
  html_document: default
  pdf_document: default
---

### Functions

Functions are a fundamental building block of R.

A function is an object in R that takes some input objects (called arguments) and returns an ouput object.

A function has a form:
```{r, eval=FALSE}
function_name <- function(argument_1, argument_2, ...) {
  expression_1
  expression_2
  ...
  return(output)
}
```
* ```argument_1, arguments_2``` are the names of variables.
* ```name``` is the name of function. You change ```name``` as you want.

To call or run the function we type:
```{r, eval=FALSE}
function_name(x1, x2, ...)
```
the value of this expression is the value of output.

* A function may have more than one ```return``` statement, in which case it stops after executing the first one it reaches.
* If there is no statement ```return(output)``` then the value returned by the function is the value of the last expression in the braces.

* A function always returns a value.
  * For some functions the value returned is unimportant.
  * In such cases one usually omits the return statement, or returns NULL.
* If the value returned by a function is not assigned to a variable, then it is printed.

* The most important advantage of using a function is that once it is loaded, it can be used again and again without having to reload it.

* The second most important use of functions is to break down a programming task into smaller logical units.

All work in R is done by functions.
For example, we have
```{r}
animals <- c("cow", "dog", "cat", "pig")
```
and you want 
```{r, eval=F}
animals[4] <- "duck"
```
This is equivalent to
```{r}
`[<-`(animals, 4, "duck")
```
In practice, we do not use the above expression, but it is useful to know that every operation in R is a function.

Other example:
```{r}
x <- 3
y <- 4
# functional form of arithmetic expression
`+`(x,y)
```
```{r}
# functional form of if-then expression
`if`(x >y, "x is larger than y", "y is larger than x")
```



#### Example : Find zeros of ```a2*x^2+a1*x+a0=0```

```{r}
quad3 <- function(a0, a1, a2) {
  if(a2 == 0 && a1 == 0 && a0 == 0) {
    roots <- NA	
  } else if (a2 == 0 && a1 == 0 ) {
	  roots <- NULL
  } else if ( a2 == 0 ) {
    roots <- -a0/a1
  } else {
    discrim <- a1^2 - 4*a2*a0
    if (discrim > 0) {
      roots <- (-a1 +c(1,-1) * sqrt(a1^2-4*a2*a0))/(2*a2)
    } else if (discrim == 0) {
      roots <- -a1/(2*a2)
    } else {
      roots <- NULL
    }
  }
  return(roots)
}
```



#### Example : n choose r

The number of ways that you can choose r things from a set of n, ignoring the order, is 
$$ \frac{n!}{r!(n-r)!}  $$
```{r}
n_factorial <-function(n) {
  n_fact <- prod(1:n)
  return(n_fact)
}
n_choose_r <- function(n, r) {
  n_ch_r <- n_factorial(n)/n_factorial(r)/n_factorial(n-r)
  return(n_ch_r)
}
```


#### Example : Winsorised mean
k-th Winsorised mean of x={x1, ?€?, xn} is defined as
$$ w_k =\frac{(k+1)x_{k+1} + x_{k+2} + \cdots + x_{n-k+1} + (k+1)x_{n-k}}{n} $$
```{r}
wmean <- function(x, k) {
	x <- sort(x)
	n <- length(x)
	x[1:k] <- x[k+1]
	x[(n-k+1):n] <- x[n-k]
	return(mean(x))
}
```


#### Exmple : Swap

swap values of ```x[1]``` and ```x[2]```
```{r, eval=FALSE}
f1  swap <- function(x){
f2    y <- x[2]
f3    x[2] <- x[1]
f4    x[1] <- y
f5    return(x)
f6  }
p1  x <- c(7, 8, 9)
p2  x[1:2] <- swap(x[1:2])
p3  x[2:3] <- swap(x[2:3])
```


### Scope and its consequences

* Argument and variables defined within a function exist only within that function.
   * If you define and use a variable x inside a function, it does not exist outside the function.
   * If variables with the same name exist inside and outside a function, then they are separate and do not interact at all.
* The variable defined outside the function can be seen inside the function (provided there is not a variable with the same name defined inside).

```{r}
test <- function(x) {
  y <- x+1
  return(y)
}
test(1)
```
```{r, eval=FALSE}
y
### Error : Object "y" not found
```
```{r}
test2 <- function(x) {
  y <- x + z
  return(y)
}
z <- 1
test2(1)
z<-2
test2(1)
```

### Optional argument and default value

* To give argument_1 the default value x1 we use argument_1 = x1 within the function definition.
* If an argument has a default then it may be omitted when calling the function, in which case the default is used.


```{r}
test3 <- function(x=1, y=1, z=1) {
  return(x*100+y*10+z)
}
test3(2,2)
test3(y=2, z=2)
```

### Vector-based programming

* Many R functions are vectorised.
* To further facilitate vector-based programming, R provides functions that enable the vectorisation of user-defined functions.
* ```apply```, ```sapply```, ```lapply```, ```tapply```, and ```mapply```.
* ```sapply(X, FUN)```
   * apply function ```FUN``` to every element of vector ```X```.
   
   
#### Example : Density of primes ?€? sapply
 
* Let $\rho(n)$ be the number of primes less than or equal to n
* Then
$$ \lim_{n\rightarrow\infty} \frac{\rho(n)\log(n)}{n} \rightarrow 1.$$

* To check this, first we define a function ```prime``` that tests if a given integer is prime.

* We then use ```sapply``` to apply ```prime``` to the vector ```2:n```.

```{r}
prime <- function(n) {
  if ( n==1 ) {
    is.prime <- FALSE
  } else if ( n==2) {
    is.prime <- TRUE
  } else {
    is.prime <- TRUE
    for ( m in 2:(n/2) ) {
			if ( n%%m == 0 ) is.prime <- FALSE
    }
  }
  return(is.prime)
}	
```

```{r}
n <- 1000
m.vec <- 2:n
primes <- sapply(m.vec, prime)
num.primes <- cumsum(primes)
plot(m.vec, num.primes/m.vec, type = "l", main = "prime density", xlab = "n", ylab = "")
lines(m.vec, 1/log(m.vec), col= "red")
```

### Recursive programing

* A recursive program is one that calls itself.
* This is useful because many algorithms are recursive in nature.

#### n factorial

```{r}
nfact2 <- function(n) {
  if (n==1) {
    cat("called nfact2(1)\n")
    return(1)
  } else {
    cat("called nfact2(", n, ")\n", sep="")
    return(n*nfact2(n-1))
  }
}
```
#### Sieve of Eratosthenes
The Sieve of Eratosthenes is an algorithm for finding all of the primes less than or equal to a given number $n$.

1. Start with the list $2,3,\cdots,n$ and $p=2$.
2. Remove from the list all elements that are multiples of $p$ (but keep $p$ itself).
3. Increase $p$ to the smallest element of the remaining list that is larger than the current $p$.
4. If $p$ is larger than $\sqrt{n}$ then stop, otherwise go back to step 2.

```{r}
primesieve <- function(sieved, unsieved) {
  p <- unsieved[1]
  n <- unsieved[length(unsieved)]
  if ( p^2 > n ) {
    return(c(sieved, unsieved))
  } else {
    unsieved <- unsieved[unsieved %% p != 0]
    sieved <- c(sieved, p)
    return(primesieve(sieved, unsieved))
  }
}

```



