---
title: "Sophisticated data structures"
output: github_document
---

## Sophisticated data structures

* R provides sophisticated structure for the storage and manipulation of data.
    * simplify data representation, manipulation and analysis.

* dataframe
    * is like a matrix, 
    * but extended to allow for different object models in different columns

* list
    * is a general data structure that can house many kinds of R object.

* factor
    * special variable that represent categorical objects.
    
    
## Factor

* Data type for ordinal and categorical vectors.   
* The possible values of a factor are referred to as its level.   
* To create a factor, apply function factor to some vector.


```{r}
hair <- c("blonde", "black", "brown", "brown", "black", "gray", "none")
(hair <- factor(hair))
```

```{r}
levels(hair)
```

* Specify level using ```levels``` argument.


```{r}
hair <- factor(hair, levels=c("black", "blonde", "brown", "gray", "white", "none"))
table(hair)
```


## Dataframe

* Dataframe is a list of vectors (with equal length).   
* Each vector (column) is a variable in experiment.   
* Each row is a single observation.


Plot | Tree | Species | Diameter | Height
---- | ---- | ------- | -------- | ------
2 | 1 | DF | 39 | 20.5
2 | 2 | WL | 48 | 33.0
3 | 2 | GF | 52 | 30.0
... | ... | ... | ... | ...


### read.table

* Large dataframe are usually read into R from a file, using ```read.table```.    

```{r, eval=FALSE}
read.table(file, header=FALSE, sep="")
```

* ```file```
    * filename to be read, can be relative or absolute.   
    * the same number of values in each row.    
    * the values may be different modes, but the pattern of modes must be the same in each row.   
* ```header```
    * whether or not the first line of the file is the variable names    
* ```sep```
    * gives the character used to separate values in each row.    
    * default ```""``` has the special interpretation that a variable amount of white space (spaces, tabs, returns) can separate values.

### more about read.table

* Two commonly used variants of ```read.table```    
    * ```read.csv(file)```
        * for comma-separated data    
        * equivalent to ```read.table(file, header=TRUE, sep=",")```    

    * ```read.delim(file)```
        * for tab-delimitated data
        * equivalent to ```read.table(file, header=TRUE, sep="\t")```


### Header


