---
title: "Sophisticated data structures"
output: github_document
---

## Sophisticated data structures

* R provides sophisticated structure for the storage and manipulation of data.
    * simplify data representation, manipulation and analysis.

* dataframe
    * is like a matrix, 
    * but extended to allow for different object models in different columns

* list
    * is a general data structure that can house many kinds of R object.

* factor
    * special variable that represent categorical objects.
    
    
## Factor

* Data type for ordinal and categorical vectors.   
* The possible values of a factor are referred to as its level.   
* To create a factor, apply function factor to some vector.


```{r}
hair <- c("blonde", "black", "brown", "brown", "black", "gray", "none")
(hair <- factor(hair))
```

```{r}
levels(hair)
```

* Specify level using ```levels``` argument.


```{r}
hair <- factor(hair, levels=c("black", "blonde", "brown", "gray", "white", "none"))
table(hair)
```


## Dataframe

* Dataframe is a list of vectors (with equal length).   
* Each vector (column) is a variable in experiment.   
* Each row is a single observation.


Plot | Tree | Species | Diameter | Height
---- | ---- | ------- | -------- | ------
2 | 1 | DF | 39 | 20.5
2 | 2 | WL | 48 | 33.0
3 | 2 | GF | 52 | 30.0
... | ... | ... | ... | ...


### read.table

* Large dataframe are usually read into R from a file, using ```read.table```.    

```{r, eval=FALSE}
read.table(file, header=FALSE, sep="")
```

* ```file```
    * filename to be read, can be relative or absolute.   
    * the same number of values in each row.    
    * the values may be different modes, but the pattern of modes must be the same in each row.   
* ```header```
    * whether or not the first line of the file is the variable names    
* ```sep```
    * gives the character used to separate values in each row.    
    * default ```""``` has the special interpretation that a variable amount of white space (spaces, tabs, returns) can separate values.

### more about read.table

* Two commonly used variants of ```read.table```    
    * ```read.csv(file)```
        * for comma-separated data    
        * equivalent to ```read.table(file, header=TRUE, sep=",")```    

    * ```read.delim(file)```
        * for tab-delimitated data
        * equivalent to ```read.table(file, header=TRUE, sep="\t")```


### Header

* If a header is present, it is used to name the columns of the dataframe.    
* You can assign your own column names after reading the dataframe (using the ```names``` function).    
* Or when you read it in, using the ```col.names``` argument,
    * which should be assigned a character vector the same lengths of the column number.
* If there is no head and no ```col.names``` argument, then R uses the names "V1", "V2", etc.

### read csv file

* ufc.csv is comma separated and there is a header line.

```{r, eval=FALSE}
ufc <- read.csv("the path of ufc.csv")
```

```{r, echo=FALSE}
ufc <- read.csv("./data/ufc.csv")
```

* head and tail to examine the object

```{r}
head(ufc)
```

```{r}
tail(ufc)
```

* Each column has a unique name and we can extract that variable by means of names using ```$```.

```{r}
x <- ufc$height.m
x[1:5]
```

### accessing element in data.frame

* We can use ```[[?]]``` to extract columns.
    * ```ufc$height.m```,  ```ufc[[5]]``` and ```ufc[["height.m"]]``` are all equivalent.
    

* You can extract the elements of a dataframe directly using matrix indexing.

```{r}
ufc[1:5, 5]
```

* To select more than one of the variables in a dataframe, we use ```[?]```.
    * ```ufc[4:5]``` is equivalent to ```ufc[c("dbh.cm", "height.m")]```.

```{r}
diam.height <- ufc[4:5]    
diam.height[1:5, ]
```
```{r}
is.data.frame(diam.height)
```

* Selectin a column using ```[[?]]``` preserves the mode of the object that is being extracted.

* Whereas, ```[?]``` keeps the mode of the object from which the extraction is being made.

```{r}
mode(ufc)
```

```{r}
mode(ufc[5])
```

```{r}
mode(ufc[[5]])
```

http://r4ds.had.co.nz/vectors.html

### creating data.frame and adding column

* construct a dataframe from a collection of vectors or dataframes

```{r, eval=FALSE}
data.frame(col1=x1, col2=x2, ¡¦, df1, df2, ¡¦)
```

* ```col1``` and ```col2``` are column names

* ```x1``` and ```x2``` are vectors of equal lengths

* ```df1``` and ```df2``` are dataframes whose columns must be the same length with ```x1```, ```x2```

* create a new variable within a dataframe, by naming it and assigning it a value.

```{r}
ufc$volume.m3 <- pi*(ufc$dbh.cm/200)^2*ufc$height/2
mean(ufc$volume.m3)
```

### name of column

* ```name(df)``` return the names of the dataframe ```df```.

* To change the names of ```df``` you pass a vector of strings to ```names(df)```.

```{r}
(ufc.names <- names(ufc))
```

```{r}
names(ufc) <- c("P", "T", "S", "D", "H", "V")
names(ufc)
```

```{r}
names(ufc) <- ufc.names
```

### subset

* ```subset``` is a convenient tool for selecting rows of dataframe.
    * ```x %in% y``` returns a logical vector whose i-th element is ```TRUE``` if ```x[i]``` is in ```y```.


```{r}
fir.height <- subset(ufc, subset=species %in% c("DF", "GF"), select = c(plot, tree, height.m))
head(fir.height)
```

### write a dataframe

```{r, eval=FALSE}
wirte.table(x, file="", append=FALSE, sep=" ", row.names=TRUE, col.names=TRUE)
```

* ```x``` is a dataframe to be written.    
* ```file``` is the address of the file to write to.    
* ```append``` indicates whether or not to append.    
* ```sep``` is the character used to separate the values.    
* ```row.name``` indicates whether or not to include the row names as the first column.    
* ```col.names``` indicates whether or not to include the column names as the first row.   

## List

* We have seen that a vector is an indexed set of objects.

* All the elements of a vector to be the same type ??? numeric, character or logical ??? which is called the mode of the vector.

* List is an indexed set of objects, but the element of a list can be of different type, including other list.
    * The mode of a list is list.

* A list is created using the ```list(¡¦)``` command, with comma separated arguments.

* Single square brackets ```[ ]``` are used to select a sublist.

* Double square brackets ```[[ ]]``` are used to extract a single element.

### Example

```{r}
(my.list <- list("one", TRUE, 3, c("f", "o", "u", "r")))
```

```{r}
my.list[[2]]
```

```{r}
mode(my.list[[2]])
```

```{r}
my.list[2]
```

```{r}
mode(my.list[2])
```

```{r}
my.list[[4]][1]
```

```{r}
my.list[4][1]
```

