---
title: "R Basic Programming"
output: github_document
---



### Branching with if


if ( logical_expression ) {   
&nbsp;&nbsp;&nbsp;&nbsp;expression_1	
&nbsp;&nbsp;&nbsp;&nbsp;...   
}  



if (logical_expression) {     
&nbsp;&nbsp;&nbsp;&nbsp;expression_1   	
&nbsp;&nbsp;&nbsp;&nbsp;...   
} else {   	
&nbsp;&nbsp;&nbsp;&nbsp;expression_2 	
&nbsp;&nbsp;&nbsp;&nbsp;...  
}  

If logical_expression is TRUE then the first group of expression is executed and the second group of expression is not executed.

Conversely if logical_expression is FALSE then only the second group expression is executed.


If you type:   


if (logical_expression) {	 
&nbsp;&nbsp;&nbsp;&nbsp;expression_1  
&nbsp;&nbsp;&nbsp;&nbsp;...}  
else {  
&nbsp;&nbsp;&nbsp;&nbsp;expression_2  
&nbsp;&nbsp;&nbsp;&nbsp;...}  

then you get an error.

This is because R is believes the if statement is finished before it sees the else part.  
That is, R treats the else as the start of a new command, but there is no command that starts with an else.

The followings are equivalent.



if (logical_expression_1){    
&nbsp;&nbsp;&nbsp;&nbsp;expression_1    
} else {   
&nbsp;&nbsp;&nbsp;&nbsp;if (logical_expression_2){   
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;expression_2   
&nbsp;&nbsp;&nbsp;&nbsp;} else {   
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;expression_3   
&nbsp;&nbsp;&nbsp;&nbsp;}      
}


if (logical_expression_1){   
&nbsp;&nbsp;&nbsp;&nbsp;expression_1   
} elseif (logical_expression_2){   
&nbsp;&nbsp;&nbsp;&nbsp;expression_2   
} else {   
&nbsp;&nbsp;&nbsp;&nbsp;expression_3   
}



Example : root of quadratic
```{r}
rm(list=ls())

a0 <- 1; a1 <- 6; a2 <- 7
discrim <- a1^2 - 4*a2*a0

if (discrim > 0) {
  roots <- c( (-a1 + sqrt(a1^2 - 4*a2*a0))/(2*a2), (-a1 - sqrt(a1^2 - 4*a2*a0))/(2*a2) )
} else if (discrim == 0) {
  roots <- -a1/(2*a2)
} else {
  roots <- c() 
}

show(roots)
```

### Looping with for


for (x in vector) {  
&nbsp;&nbsp;&nbsp;&nbsp;expression_1   
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;...   
}


where x is a simple variable.

Example : summing a vector

We have a built-in function sum() but for an illustrative purpose:

```{r}
(x_list <- seq(1, 9, by = 2))
sum_x <- 0
for (x in x_list) {
  sum_x <- sum_x + x
  cat("The current loop element is", x, "\n")
  cat("The cumulative total is", sum_x, "\n")
}
```
* cat() : for concatenate
* \n : new line

