---
title: "R Basic Programming"
output: html_notebook
---



### Branching with if

```R
if (logical_expression) {  
    expression_1	
    …  
}
```

```R
if (logical_expression) {  
    expression_1	
    …  
} else {	
    expression_2 	
    … 
}
```

If logical_expression is TRUE then the first group of expression is executed and the second group of expression is not executed.

Conversely if logical_expression is FALSE then only the second group expression is executed.


If you type  

```R
if (logical_expression) {	 
  expression_1  
  …}  
else {  
  expression_2  
  …}  
```
then you get an error.

This is because R is believes the if statement is finished before it sees the else part.  
That is, R treats the else as the start of a new command, but there is no command that starts with an else.

The followings are equivalent.

```{r}

if (logical_expression_1){
  expression_1 
} else {
	  if (logical_expression_2){
	    expression_2
	 } else {
		  expression_3
	}
}
```
```{r}
if (logical_expression_1){
  expression_1
}  else {
	if (logical_expression_2){
	  expression_2
  } else {
		expression_3
	}
}

```

Example : root of quadratic
```{r}
rm(list=ls())

a0 <- 1; a1 <- 6; a2 <- 7
discrim <- a1^2 - 4*a2*a0

if (discrim > 0) {
  roots <- c( (-a1 + sqrt(a1^2 - 4*a2*a0))/(2*a2), (-a1 - sqrt(a1^2 - 4*a2*a0))/(2*a2) )
} else if (discrim == 0) {
  roots <- -a1/(2*a2)
} else {
  roots <- c() 
}

show(roots)
```

### Looping with for

```{r}
for (x in vector) {
  expression_1
  …
}
```
where x is a simple variable.

Example : summing a vector


