---
title: "Programming with functions"
output: github_document
---

### Functions

A function has a form:
```{r, eval=FALSE}
name <- function(argument_1, argument_2, ...) {
  expression_1
  expression_2
  ...
  return(output)
}
```
* ```argument_1, arguments_2``` are the names of variables.
* ```name``` is the name of function. You change ```name``` as you want.

To call or run the function we type:
```{r, eval=FALSE}
name(x1, x2, …)
```
the value of this expression is the value of output.

* A function may have more than one ```return``` statement, in which case it stops after executing the first one it reaches.
* If there is no statement ```return(output)`` then the value returned by the function is the value of the last expression in the braces.

* A function always returns a value.
  * For some functions the value returned is unimportant.
  * In such cases one usually omits the return statement, or returns NULL.
* If the value returned by a function is not assigned to a variable, then it is printed.

* The most important advantage of using a function is that once it is loaded, it can be used again and again without having to reload it.

* The second most important use of functions is to break down a programming task into smaller logical units.

#### Example : Find zeros of ```a2*x^2+a1*x+a0=0```

```{r}
quad3 <- function(a0, a1, a2) {
  if(a2 == 0 && a1 == 0 && a0 == 0) {
    roots <- NA	
  } else if (a2 == 0 && a1 == 0 ) {
	  roots <- NULL
  } else if ( a2 == 0 ) {
    roots <- -a0/a1
  } else {
    discrim <- a1^2 - 4*a2*a0
    if (discrim > 0) {
      roots <- (-a1 +c(1,-1) * sqrt(a1^2-4*a2*a0))/(2*a2)
    } else if (discrim == 0) {
      roots <- -a1/(2*a2)
    } else {
      roots <- NULL
    }
  }
  return(roots)
}
```



#### Example : n choose r

The number of ways that you can choose r things from a set of n, ignoring the order, is 
$$ \frac{n!}{r!(n-r)!}  $$
```{r}
n_factorial <-function(n) {
  n_fact <- prod(1:n)
  return(n_fact)
}
n_choose_r <- function(n, r) {
  n_ch_r <- n_factorial(n)/n_factorial(r)/n_factorial(n-r)
  return(n_ch_r)
}
```


#### Example : Winsorised mean
k-th Winsorised mean of x={x1, …, xn} is defined as
$$ w_k =\frac{(k+1)x_{k+1} + x_{k+2} + \cdots + x_{n-k+1} + (k+1)x_{n-k}}{n} $$
```{r}
wmean <- function(x, k) {
	x <- sort(x)
	n <- length(x)
	x[1:k] <- x[k+1]
	x[(n-k+1):n] <- x[n-k]
	return(mean(x))
}
```


#### Exmple : Swap

swap values of ```x[1]``` and ```x[2]```
```{r, eval=FALSE}
f1  swap <- function(x){
f2    y <- x[2]
f3    x[2] <- x[1]
f4    x[1] <- y
f5    return(x)
f6  }
p1  x <- c(7, 8, 9)
p2  x[1:2] <- swap(x[1:2])
p3  x[2:3] <- swap(x[2:3])
```

![alt tag]()

### Scope and its consequences

* Argument and variables defined within a function exist only within that function.
   * If you define and use a variable x inside a function, it does not exist outside the function.
   * If variables with the same name exist inside and outside a function, then they are separate and do not interact at all.
* The variable defined outside the function can be seen inside the function (provided there is not a variable with the same name defined inside).

```{r}
test <- function(x) {
  y <- x+1
  return(y)
}
test(1)
y
```


