---
title: "1.Introduction to R"
output: 
  ioslides_presentation:
    css: custom.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## R 

* R is an implementation of a functional programming language S.
* R has been developed and maintained by a core of statistical programmers, with the support of a large community.
* R is free.
* R is most widely used for statistical computing and graphics.
* R is also a fully programming language well suited to scientific programming in general.


## R: pro and cons {.smaller}

* Pro: A picture says more than a thousands words  
     + R and visualization are a perfect match. (ggplot2, ggvis, googleVis and rCharts.)
* Pro: R ecosystem
     + R has a rich ecosystem of cutting-edge packages and active community. 
* Pro: R lingua franca of data science
     + R is developed by statisticians for statisticians who communicate ideas and concepts through R code and packages
* Pro/Con: R is slow
     + Although R can be experienced as slow due to poorly written code, there are multiple packages to improve R?۪s performance: pqR, renjin and FastR, Riposte and many more.
* Con: R has a steep learning curve

<http://www.kdnuggets.com/2015/05/r-vs-python-data-science.html>

## R as a calculating environment {.smaller}
<div class="columns-2">

R can be used as a powerful calculator.  
Arithmetic operations: 

```{r, echo = TRUE}
(1+1/100)^100
```
```{r, echo = TRUE}
17%%5
```
```{r, echo = TRUE}
17%/%5
```

symbol | meaning
-------|---------
+ | addition
- | subraction
* | multiplication
/ | division
^ | exponential
%% | modulus
%/% | integer division
</div>


## R has a number of built in functions {.smaller}

sin(x), cos(x), tan(x), exp(x), log(x), sqrt(x), floor(x), ceiling(x), round(x), ...

```{r echo=TRUE}
exp(1)
```
```{r echo=TRUE}
options(digits = 16)
exp(1)
```
```{r echo=TRUE}
pi
sin(pi/6)
```

## Variable

* We can assign a value to a variable and use the variable.
* For the assignment, we use command <-.
* Variable names made up of letters, numbers, . or _
    * provided it starts with a letter, or . then a letter.
    * names are case sensitive.
    * for example,
         * x, y, my_variable, a1, a2, .important_variable, x.input
    * wrong name:
         * 2016_income, .1grade, _x, y@gmail.com


## Variable {.smaller}
<div class="columns-2">
To display the value of a variable x, we type x 
* or print(x) or show(x).
    
```{r}
x <- 100
x
```

We can show the outcome of assignment by parentheses.

```{r}
(y <- (1+1/x)^x)
```
When assigning, the right-hand side is evaluated first, then that value is placed in the variable on the left-hand side.
```{r}
n <- 1
n <- n+1
n
```

R allows the use of = for variable assignment, in common with most programming languages.
</div>


## Functions {.smaller}

Takes one or more argument (inputs) and produces one or more outputs (return values).

```{r}
seq(from = 1, to = 9, by =2)
```
```{r}
seq(from = 1, to = 9)
```

You can access the built-in help by 
```{r, eval=FALSE}
help(function_name)
```
or 
```{r, eval=FALSE}
?function_name
```

----

Every function has a default order for arguments.  
If you provide arguments in this order, then they do not need to be named.
```{r}
seq(1, 9, 2)
```
```{r}
seq(to = 9, from = 1)
```
```{r}
seq(by=-2, 9, 1)
```

## Vectors

* indexed list of variables
* the name of the i-th element of vector x is x[i]
* three basic functions for constructing vectors

```{r}
(x <- seq(1, 20, by = 2))
```

```{r}
(y <- rep(3, 4))
```

```{r}
(z <- c(y, x))
```

* another method for sequence
```{r}
(x <- 100:110)
```
```{r}
(y <- 110:100)
```

* vector and index
```{r}
(x<- 100:110)

i <- c(1, 3, 2)

x[i]
```

minus index:
```{r}
j <- c(-1, -2, -3)
x[j]
```

empty vetor:
```{r}
x <- c()
```

elementwise algebraic operation:
```{r}
x <- c(1, 2, 3)
y <- c(4, 5, 6)
x*y
y^x
```

with unequal length of vectors:
```{r}
c(1, 2, 3, 4) + c(1, 2)

2 * c(1, 2, 3)

(1:3)^2
```

This works but with warning message:
```{r}
c(1, 2, 3) + c(1, 2)
```

functions taking vectors
```{r}
sqrt(1:3)

mean(1:6)

sort(c(5, 1, 3, 4, 2))
```

Example : mean and variance

compare computed mean and variance with built-in functions

```{r}
x <- c(1.2, 0.9, 0.8, 1, 1.2)

x.mean <- sum(x)/length(x)

x.mean - mean(x) 
```
```{r}
x.var <- sum((x-x.mean)^2)/(length(x)-1)
x.var - var(x)
```


Example : simple numerical integration

```{r}
dt <- 0.005
t <- seq(0, pi/6, by = dt)
ft <- cos(t)
(I <- sum(ft) * dt)
```
t is a vector and ft is also a vector.
```{r}
I - sin(pi/6)
```
Note the difference between the numerical integration and theoretical value.


Example : exponential limit
```{r}
x <- seq(10, 200, by = 10)

y <- (1 + 1/x)^x

exp(1) - y
```
```{r}
plot(x, y)
```


## Matrix example {.smaller}


```{r}
(A <- matrix(1:6, nrow=2, ncol=3, byrow=TRUE))
```
```{r}
A[1, 3] <- 0
A
```
```{r}
A[, 2:3]
```

---- 

```{r}
(B <- diag(c(1,2,3)))
```


